A formal treatment of bidirectional typing\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}

### 2023-12-21 tex/ESOP24/S3-defs.tex
[ tex ] Minor tweaks
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

\section{Bidirectionally simply typed languages}\label{sec:defs}
This section provides general definitions
%\footnote{%
%Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
of simple types, simply typed languages, and bidirectional type systems, and uses simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
These definitions may look dense, especially on first reading.
The reader may choose to skim through this section, in particular the figures, and still get some rough ideas from later sections.

The definitions are formulated in two steps:
\begin{inlineenum}
  \item first we introduce a notion of arity and a notion of signature which includes a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Also, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory.
  } of operation symbols and an assignment of arities to symbols;
\item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
\end{inlineenum}
%All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.

@@

\begin{definition} \label{def:substitution}
The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
\end{definition}

@@

The set~$\Xi$ of type variables for each operation, called its \emph{local context}, plays a somewhat important role:
To use a rule like $\Rule{Abs}$ in an actual typing derivation, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
In our formulation of substitution~(\cref{def:substitution}), we must first identify for which type variables to substitute.
As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include functions from~$\Xi$ to concrete types specifying how to instantiate typing rules by substitution.

@@

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{$\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a language $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
    \and
    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
    \\
  \text{for $\bop$ in $\Omega$}
  \end{mathpar}
  \caption{Typing derivations}
  \label{fig:extrinsic-typing}
\end{figure}
\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional system $(\Sigma, \Omega)$} 
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
    \\ \text{for $\biop$ in $\Omega$}
  \end{mathpar}
  \caption{Bidirectional typing derivations}
  \label{fig:bidirectional-typing-derivations}
\end{figure}

>>>>

\section{Bidirectionally simply typed languages}\label{sec:defs}
This section provides general definitions
%\footnote{%
%Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
of simple types, simply typed languages, and bidirectional type systems, and uses the simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
These definitions may look dense, especially on first reading.
The reader may choose to skim through this section, in particular the figures, and still get some rough ideas from later sections.

The definitions are formulated in two steps:
\begin{inlineenum}
  \item first we introduce a notion of arity and a notion of signature which includes a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Indeed, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory~\citep[Definition 3.1.1]{UFP2013}.} of operation symbols and an assignment of arities to symbols;
\item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
\end{inlineenum}
%All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.

@@

\begin{definition}\label{def:substitution}
The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
\end{definition}

@@

The set~$\Xi$ of type variables for each operation, called its \emph{local context}, plays a somewhat important role:
To use a rule like $\Rule{Abs}$ in an actual typing derivation, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
In our formulation of substitution~(\ref{def:substitution}), we must first identify which type variables to substitute for.
As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include functions from~$\Xi$ to concrete types specifying how to instantiate typing rules by substitution.

@@

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{$\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a language $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
    \and
    \inferrule{\rho : \Sub{\Xi}{\emptyset} \and  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \and \cdots\and \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
    \\
  \text{for $\bop$ in $\Omega$}
  \end{mathpar}
  \caption{Typing derivations}
  \label{fig:extrinsic-typing}
\end{figure}
\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional system $(\Sigma, \Omega)$} 
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \and \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \and
      \cdots \and
    \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\Rule{Op}
    \\ \text{for $\biop$ in $\Omega$}
  \end{mathpar}
  \caption{Bidirectional typing derivations}
  \label{fig:bidirectional-typing-derivations}
\end{figure}

### 2023-12-21 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
We also discuss some related properties:
By bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
We also formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.

@@

The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{For the sake of simplicity, we use ordinary mode decoration elsewhere in this paper.}

@@

This lemma would be wrong if the `strictness' boolean~$s$ was left out of the rules:
Having both $\ChkRule{Sub}$ and $\SynRule{Missing}$, which we call \emph{mode casts}, it would be possible to switch between the two modes freely, which unfortunately means that we could insert a pair of $\ChkRule{Sub}$ and $\SynRule{Missing}$ anywhere, constructing a non-genuine derivation even when there is in fact a genuine one.
The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts:
Every rule other than the mode casts sets~$s$ to~$\BTrue$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\BFalse$.
Furthermore, the sub-derivation of a mode cast has to be strict, so it is impossible to have consecutive mode casts.
Another way to understand the role of~$s$ is that it makes the $\SynRule{Missing}$ rule precise: an annotation is truly missing only when a term is \emph{strictly} in checking mode but is required to be in synthesising mode.
With strictness coming into play, non-genuine derivations are now `truly non-genuine'.

@@

\begin{proposition}[Annotatability]\label{thm:annotatability}
If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.

>>>>

Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
We also discuss some related properties: by bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
We also formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.

@@

The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better: if a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{For the sake of simplicity, we use ordinary mode decoration elsewhere in this paper.}

@@

This lemma would be wrong if the `strictness' boolean~$s$ was left out of the rules: having both $\ChkRule{Sub}$ and $\SynRule{Missing}$, which we call \emph{mode casts}, it would be possible to switch between the two modes freely, which unfortunately means that we could insert a pair of $\ChkRule{Sub}$ and $\SynRule{Missing}$ anywhere, constructing a non-genuine derivation even when there is in fact a genuine one.
The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts: every rule other than the mode casts sets~$s$ to~$\BTrue$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\BFalse$.
Furthermore, the sub-derivation of a mode cast has to be strict, so it is impossible to have consecutive mode casts.
Another way to understand the role of~$s$ is that it makes the $\SynRule{Missing}$ rule precise: an annotation is truly missing only when a term is \emph{strictly} in checking mode but is required to be in synthesising mode.
With strictness coming into play, non-genuine derivations are now `truly non-genuine'.

@@

\begin{proposition}[Annotatability]\label{thm:annotatability}
If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}
\LT{the paper is perhaps too modest: (...) I don't think DK annotatability says anything about evidence---a synthesiser that answered "no" could still enjoy annotatability (even if we would not enjoy that answer).}
Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.

### 2023-12-21 tex/ESOP24/S5-synthesis.tex
[ tex ] Update Future work
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction on the typing derivations.
However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
\begin{lemma}[Uniqueness of synthesised types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true.
        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
      \end{enumerate}
  \end{itemize}
\end{proof}

@@

Combining the bidirectional type synthesiser with the mode decorator, soundness, and completeness from \cref{sec:pre-synthesis}, we get a type synthesiser parameterised by $(\Sigma, \Omega)$, generalising \cref{thm:implementation}.

\begin{corollary}[Trichotomy on raw terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$,
    \item $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$, or 
    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}

>>>>

Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction on the typing derivations.
However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
\begin{lemma}[Uniqueness of synthesised types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true.
        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
      \end{enumerate}
  \end{itemize}\qed
\end{proof}

@@

Combining the bidirectional type synthesiser with the mode decorator, soundness, and completeness from \cref{sec:pre-synthesis}, we derive a type synthesiser parameterised by~$(\Sigma, \Omega)$, generalising \cref{thm:implementation}.

\begin{corollary}[Trichotomy on raw terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and $\Gamma |-_{\Sigma, \erase{\Omega}} \isTerm{t} : A$ for some type~$A$,
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but $\Gamma |/-_{\Sigma, \erase{\Omega}} \isTerm{t} : A$ for any type~$A$, or 
    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}

### 2023-12-21 tex/ESOP24/S8-future.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule} which is not syntax-directed:
\[
  \inferrule{\Gamma \vdash t \syn \forall \alpha. A}{\Gamma \vdash t \syn A[B/\alpha]}
\]
By translating the above rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A general theory that accommodates various solutions to the problem is left as future work.

@@

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They proposes a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.), a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of generic bidirectional type synthesis which is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the set of computational rules is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.

>>>>

However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}:
\[
  \inferrule{\Gamma \vdash t \syn \forall \alpha. A}{\Gamma \vdash t \syn A[B/\alpha]}
\]
which is not syntax-directed.
By translating the above rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A theory that accommodates various solutions to the problem is left as future work.

@@

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.), a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of generic bidirectional type synthesis which is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the set of computational rules is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.
